---
title: "Economic and health consequences of storm events in the US"
author: "Coursera Reproducible Research Assignment 2"
output: html_document
---

# Synopsis

In this report we aim to describe which types of events are most harmful with respect to population health and have the greatest economic consequences in the United States between the years 1950 et 2011.
Our overall hypothesis is that danger to humans should come from extreme conditions, while economic consequences will be strongest where water is involved.
To investigate this hypothesis, we obtained the NOAA Storm Database from the National Oceanic
 and Atmospheric Administration (NOAA).
We specifically obtained data for the years 1950 to november 2011 (the most recent data available). 
From these data, we found that tornadoes are by far the most dangerous events for human health, while floods generate the biggest costs.


# Data Processing

## Reading in the  data

We first read in the provided data from the csv file included in the .bz2 archive. The data is a comma delimited file were fields are separated with the , character and missing values are coded as blank fields. String fields are further delimited with a set of "" characters to make sure commas inside a field is not mistakenly taken as a separator.

The first line contains the titles which we read as they are, which is the default behaviour of the `read.csv` function. We will consider cleaning the names after we create a subset with the columns we need.

```{r loading, cache=TRUE}
if (!file.exists("repdata_data_StormData.csv.bz2"))
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",destfile = "repdata_data_StormData.csv.bz2",method="curl")
storms <- read.csv("repdata_data_StormData.csv.bz2")
```

## Exploratory data analysis

After reading the data we check the basic information

```{r}
dim(storms)
names(storms)
```
We can see we have 902297 rows and 37 columns.

The first column we are interested in is the `EVTYPE` column. According to the documentation (section 2.1.1,  Storm Data Event Table) we should have 48 different values. In order to check if the data is clean in that order, we create a vector with each possible values, and see if the data in the `EVTYPE` column match those values.

```{r}
events <- c("Astronomical Low Tide","Hurricane (Typhoon)","Avalanche","Ice Storm","Blizzard","Lake-Effect Snow","Coastal Flood","Lakeshore Flood","Cold/Wind Chill","Lightning","Debris Flow","Marine Hail","Dense Fog","Marine High Wind","Dense Smoke","Marine Strong Wind","Drought","Marine Thunderstorm Wind","Dust Devil","Rip Current","Dust Storm","Seiche","Excessive Heat","Sleet","Extreme Cold/Wind Chill","Storm Surge/Tide","Flash Flood","Strong Wind","Flood","Thunderstorm Wind","Frost/Freeze","Tornado","Funnel Cloud","Tropical Depression","Freezing Fog","Tropical Storm","Hail","Tsunami","Heat","Volcanic Ash","Heavy Rain","Waterspout","Heavy Snow","Wildfire","High Surf","Winter Storm","High Wind","Winter Weather")
 table(storms$EVTYPE %in% events)
```
The results are so low that we must be missing something. We try again with a case-insensitive  comparison
```{r}
table(tolower(storms$EVTYPE) %in% tolower(events))
```
We see that around 30% of event types are not normalized. Let's see which one are more common.

```{r}
library(dplyr)
notNormalizedCount <- storms %>%
    filter(!(tolower(EVTYPE) %in% tolower(events))) %>%
    group_by(EVTYPE) %>% 
    summarise(total=n()) %>%
    arrange(-total)
head(notNormalizedCount,n=10)
```

We will try to reassign the normalized type where it is most significant, meaning where high casulaties or damage where reported. The next step is thus to clean the data.

## Data Cleaning

First  we create a new `propcost` column for property damages by calculating the actual numeric  dollar amount from the `PROPDMG` and `PROPDMGEXP` columns. We assign 0 when the multiplier is neither K, M or B.

```{r}
storms= storms %>% 
    mutate(propcost=ifelse(PROPDMGEXP=='K',as.numeric(PROPDMG*1000),
                           ifelse(PROPDMGEXP=='M',as.numeric(PROPDMG*1000000),
                                  ifelse(PROPDMGEXP=='B',as.numeric(PROPDMG*1000000000),
                                         0))))
```

We do the same operation for damages to crops, by creating a `cropcost` column from the `CROPDMG` and `CROPDMGEXP` columns

```{r}
storms= storms %>% 
    mutate(cropcost=ifelse(CROPDMGEXP=='K',as.numeric(CROPDMG*1000),
                           ifelse(CROPDMGEXP=='M',as.numeric(CROPDMG*1000000),
                                  ifelse(CROPDMGEXP=='B',as.numeric(CROPDMG*1000000000),
                                         0))))
```

Finally we create a totalcost column for each event, which is the sum of the property and crop damages economic cost.
```{r}
storms <- storms  %>%  mutate(ecocost=propcost+cropcost)
```

The second part of data cleaning is taking into account the  harmfulness to population health. For this, we just sum the number of fatalities and injuries for each event

```{r}
storms <- storms  %>%  mutate(harmfulness=FATALITIES+INJURIES)
```


The last part is to normalize the significant events, as mentionned in our Exploratory Data Analysis. As an example we choose to normalize all event types that have the 5 most occurences

```{r}
# display and normalize the 5 most common non-normalized occurrences
head(notNormalizedCount,n=5)
storms = storms %>% mutate(type=ifelse(tolower(storms$EVTYPE) %in% tolower(events),as.character(EVTYPE),
                              ifelse(EVTYPE=='TSTM WIND','THUNDERSTORM WIND',
                                     ifelse(EVTYPE=='THUNDERSTORM WINDS','THUNDERSTORM WIND',
                                            ifelse(EVTYPE=='MARINE TSTM WIND','THUNDERSTORM WIND',
                                                   ifelse(EVTYPE=='URBAN/SML STREAM FLD','FLASH FLOOD',
                                                          ifelse(EVTYPE=='HIGH WINDS','HIGH WIND',
                                                                 as.character(EVTYPE)))))))) 

```

Finally, we only keep the columns we need for the final result computations

```{r}
storms = storms %>%
    transmute(type=type,
              harmfulness=harmfulness,
              ecocost=ecocost)
```

# Results

We first report the 5 events most harmful to health

```{r}
library(ggplot2)

harmCount <- storms %>%
    group_by(type) %>% 
    summarise(total=sum(harmfulness)) %>%
    arrange(-total)

head(harmCount,n=5)

ggplot(head(harmCount,n=5), aes(x=type, y=total)) + 
  geom_bar(stat="identity")+
  labs(x="event",y="count",title="5 most harmful events")

```

We then show the 5 most economically significant events

```{r}
ecoCount <- storms %>%
    group_by(type) %>% 
    summarise(total=sum(ecocost)) %>%
    arrange(-total)
head(ecoCount,n=5)
ggplot(head(ecoCount,n=5), aes(x=type, y=total)) + geom_bar(stat="identity")+
    labs(x="event",y="cost ($)",title="5 most costly events")


```

We can see that tornadoes are by far the most dangerous events for human health, while floods generate the biggest costs.

